Alpine snow algae species composition along an elevational gradient

### 2020-01-13 manuscript/casey1.Rmd
draft to lab, made krienitzii a small supplemental

---
title: "Alpine snow algae blooms vary with elevation"
author:
  - name: Casey B. Engstrom
    institute: SFU
  - name: Kurt M. Yakimovich
    institute: SFU
  - name: Lynne M. Quarmby
    institute: SFU
    email: quarmby@sfu.ca
institute:
  - SFU: Dept. Molecular Biology and Biochemistry, Simon Fraser University, 8888 University Drive, Burnaby, BC, V5A 1S6, Canada
output: 
  bookdown::word_document2:
    fig_caption: yes
    pandoc_args:
      - '--lua-filter=templates/scholarly-metadata.lua'
      - '--lua-filter=templates/author-info-blocks.lua'
    reference_docx: templates/word-template.docx
  # pdf_document:
  #   fig_caption: yes
  #   number_sections: yes
csl: templates/the-isme-journal.csl
bibliography: refs.bib

---

@@

@@

Snow algae blooms cover vast areas of summer snowfields worldwide, and can reduce albedo and increase snow melt. Despite their global prevalence, little is known about the algae species that comprise snow algae blooms. We used 18S and *rbcL* metabarcoding and light microscopy to characterise species composition of 33 snow algae samples from alpine and subalpine habitats in the Coast Range of British Columbia, Canada. We found blooms were dominated by the snow algae genera *Sanguina*, *Chloromonas*, and *Chlainomonas*. Above treeline blooms contained *Sanguina*, *Raphidonema*, and one OTU of *Chloromonas*, distinct from species assemblages below treeline predominated by *Chloromonas krienitzii*. Previously unrecognized as an important component of the snow algae microbiome, *Chlainomonas* were abundant in samples across all elevations. Our *rbcL* dataset detected diversity that was obscured by the low taxonomic resolution of traditional 18S metabarcoding, including OTUs that could represent novel species of snow algae. These three cross-referenced datasets show alpine snow algae blooms are more diverse than previously thought, and are highly variable from site to site along an elevational gradient.

@@

# Introduction {-}
<!-- global, vast blooms. albedo->snowmelt. what species? -->
Each summer in polar and alpine snowfields worldwide, vast areas of snow surface are colored red by snow algae blooms. Such blooms are reported from every continent [@segawa_bipolar_2018;@duval_snow_1999;@marchant_snow_1982;@yoshimura_community_1997;@vimercati_nieves_2019] and Arctic sea ice [@gradinger_snow_1996]. They can be extensive: in Alaska, snow algae spectral signatures were detected in one third of a 1,900 km^2^ icefield [@ganey_role_2017]. In recent years snow algae have received attention for their role in reducing snow surface albedo, which could substantially impact snow melt rates [@lutz_biogeography_2016;@ganey_role_2017]. Thus, snow algae could impact the timing of spring melt and runoff, depleting summer water supplies held in mountain snowpacks and reducing glacier mass balance. Despite their global ubiquity, we are only beginning to identify the species that cause snow algae blooms.

<!-- why microscopy not reliable -->
Much of the previous work on snow algae relied on cell morphology for identification, but microscopy-based identification of snow algae can be unreliable. Different species of snow algae can look nearly identical: recent phylogenetic work found that four distinct field samples dominated by similar-looking cells previously referred to as *Chloromonas cf. nivalis* are actually genetically distinct clades [@matsuzaki_taxonomic_2019]. Furthermore, the same species can look completely different in different environmental conditions or life stages. For example, cultured *Chloromonas krienitzii* cells are small green biflagellates, but field samples cells of this species are larger orange spheres with thick walls and short spines [@matsuzaki_revisiting_2015]. In the freshwater green algae *Haematococcus pluvialis* red secondary pigments are produced in response to high light [@shah_astaxanthin-producing_2016], but genetic comparison of red and green snow found they contained different species [@terashima_microbial_2017;@lutz_integrated_2015].

@@

<!-- metabarcoding -->
In contrast to the high diversity of snow algae species observed in culture, metabarcoding suggests that snow algae blooms are highly similar across continents, dominated by few taxa. One study of 33 red snow communities across the Arctic found all samples were dominated by two OTUs of "uncultured Chlamydomonadaceae 1 and 2" with low relative abundance of *Chloromonas polyptera* and *Raphidonema nivale* [@lutz_biogeography_2016]. Similarly, another study using ITS2 metabarcoding found 24 red snow sites across the Arctic and Antarctic contained similar assemblages, dominated by "uncultured Chlamydomonadaceae A and B" with secondary abundance of *Raphidonema* and *Chloromonadinia* [@segawa_bipolar_2018]. Other 18S metabarcoding studies of snow algae were limited to genus or family level taxonomy [@hamilton_primary_2017]---because 18S is highly conserved, this marker generally cannot distinguish between closely related taxa.

<!-- our goal, what we did, what we found -->
Virtually nothing is known about the regional variation in species composition of snow algae blooms, particularly in alpine regions. We set out to identify the algae in blooms on different mountains and elevations in the Coast Range of British Columbia, Canada. We used 18S and *rbcL* metabarcoding to assess species composition---the latter targeting a hypervariable region of *rbcL* (coding for large subunit of Rubisco), thus targeting only photosynthetic species with higher taxonomic resolution than 18S. We also used light microscopy to describe the relative abundance of different morphologies in each field sample. Using these three cross-referenced metrics we described the algal species assemblage in our study area with high accuracy. Using *rbcL* as a marker revealed much greater diversity than would have been found using 18S alone. We did not find any differences in bloom composition between mountains, but there were distinct algae species assemblages at high versus low elevations.

@@

<!-- refer to each supp at least once in txt -->
<!-- add cities, states -->

@@

<!-- the problem with this is we cherry picked our samples. We weren't doing transects or anything like that. A reviewer could say, you couldve picked your samples that supported your hypothesis about different bloom types. which is why I brought in the cell count data. -->

We collected snow algae from sites throughout the Coast Range near Vancouver, British Columbia, Canada over the summer of 2018 (Supplementary Figure S1). To capture the extent of snow algae diversity we sampled from different elevations, dates, mountains, and micro-habitats within the snow. In total, we collected 309 colored snow samples from a range of 880 - 2150 m elevation, including samples from 13 different mountains collected on 33 different dates (Supplementary Table S1). Our samples trended upwards in elevation over the course of the summer as the snow melted earliest at the lowest elevations. We first detected snow algae above 1500 m on June 20. We scooped samples from visibly colored snow into sterile 50 mL centrifuge tubes. To prevent melting en route to the lab we stored tubes in a bag of snow. We melted each sample at room temperature on the bench and removed a 1 mL aliquot for light microscopy. Immediately after, samples were stored at -20 &deg;C for up to eight months until DNA extraction.

We used light microscopy to characterize the cell morphologies in each sample within 24 hours following collection. We fixed samples in 2% gluteraldehyde, and under 400x magnification observed the 100 cells closest to the center of the slide. We classified cell morphology based on similarity to published photographs of *Sanguina nivaloides* [@prochazkova_sanguina_2019], *Chloromonas cf. nivalis* [@prochazkova_chloromonas_2018], *Chloromonas cf. brevispina* [@matsuzaki_revisiting_2015], *Chloromonas krienitzii* [@matsuzaki_revisiting_2015], *Chlainomonas rubra* [@novis_two_2008], and for cells that did not fall into one of there categories as either "green cell" or "other".

<!-- how do select field and micro photos fit into this? -->

@@

<!-- This part is non reproducible, selecting based on date, etc -->

We selected 33 out of our 310 samples for *rbcL* and 18S Illumina amplicon sequencing. We chose these samples to represent the variation in date, elevation, geographic location, snow color, micro-habitat, and cell morphology. We freeze-dried these samples for up to 48 hours until until samples appeared completely desiccated. We mini-pestled between 5 to 20 mg of sample at room temperature to physically rupture cell walls before incubation in CTAB lysis buffer [@noauthor_ctab_2009]. We extracted DNA using chloroform:isoamyl alcohol [@cubero_dna_1999], and spin columns (Qiagen, Hilden) (details available in Supplementary Protocol S1). As a negative control we processed a sterile distilled water sample alongside each batch.

We designed custom primers to target a hypervariable region of snow algae *rbcL*. *rbcL* is more differentiated between snow algae than 18S (Supplementary Figure S2), and contains similar number of reference sequences on GenBank. Our *rbcL* primers target a 400 bp section of snow algae *rbcL* based off the consensus of 20 GenBank snow algae sequences (Supplementary Figure S2). We designed our primers with the Eurofins primer design tool (https://www.eurofinsgenomics.eu/en/ecom/tools/pcr-primer-design/). *Sanguina* sequences were not included because they were not available at the time. To target 18S, we used the universal primers Euk1181 and Euk1624 targeting the V7-V8 hypervariable regions [@wang_optimal_2014]. Primer sequences are available in Supplementary Table S2.

We constructed our 18S and *rbcL* amplicon libraries using a standard two-step PCR approach [@meyer_illumina_2010]. For detailed PCR conditions, see Supplementary Protocol S2. After each PCR, we purified using Agencourt AMPure XP kit (Beckman Coulter, Brea, CA). We standardized sample DNA concentration using Qubit fluorometric quantification (Thermo Fisher, Waltham, MA). The pooled library was loaded and run on a MiSeq V3 kit (Illumina, San Diego, CA).

@@

We demultiplexed reads using CUTADAPT [@martin_cutadapt_2011], and filtered and trimmed, removed errors, dereplicated, merged paired-end reads, and removed chimeras following the default pipeline of DADA2 [@callahan_dada2_2016]. We assigned taxonomy to amplicon sequence variants (ASVs) using IDTaxa [@murali_idtaxa_2018], only assigning names to ASVs that were assigned with high confidence (threshold set to 50). We assigned taxonomy using custom snow algae reference databases for both 18S and *rbcL*, in addition to re-assigning 18S with SILVA to classify non-snow algae [@quast_silva_2013]. Snow algae reference databases and the scripts to generate them are available on our github repository.

@@

<!-- make sure this is on GH: -->

@@

To generate OTUs we clustered ASVs using t-SNE [@maaten_visualizing_2008]. We compared t-SNE results with the perplexity parameter ranging from 1 to 100, and found comparable results from 10 to 50. We categorized OTUs by applying DBSCAN [@hahsler_dbscan_2019] to our t-SNE. We compared *rbcL* sample similarity using UniFrac with default settings [@lozupone_unifrac_2005]. To confirm that our clustering was not an artifact of t-SNE we created maximum likelihood phylogenetic trees of ASVs using IQTree [@nguyen_iq-tree:_2015] (Supplementary Figure S3).

# Results {-}
<!-- should we report the diversity data quickly? or tangential to our question... -->

<!-- add read depth, etc? -->
<!-- make fig 5, ???, and make 18s otus -->

We observed morphologically distinct snow algae communities at different elevations (Figure 1; Supplementary Figure S4). Red snow was prevalent in areas of high solar exposure above treeline, predominantly containing cells similar in appearance to *Sanguina nivaloides* [@prochazkova_sanguina_2019]. We occasionally observed green snow patches in alpine meadows receiving full sun, but more frequently observed green snow in well-shaded, forested sites. We found cells resembling *Chlainomonas rubra* [@novis_two_2008] at all elevations. Forested sites at low elevations often contained green or orange patches of snow, predominated by green or orange ovate cells with short spines resembling *Chloromonas cf. brevispina* [@matsuzaki_revisiting_2015].

Both 18S and *rbcL* amplicon libraries were dominated by ASVs that were assigned to Chlorophyta. Our 18S library contained 67 ASVs that were assigned to Chlorophyta and 7 ASVs to Ochrophyta, while our *rbcL* library contained 603 ASVs assigned to Chlorophyta and 41 to Trebouxiophyceae. Negative controls did not contain Qubit-detectable amounts of DNA. Our *rbcL* library contained higher ASV and OTU diversity than 18S, t-SNE clustering generated 9 *rbcL* OTUs, while we detected just 3 visually distinct clusters among 18S ASVs (Supplementary Figure S5). For *rbcL*, the most frequent taxonomic assignments at genus level were *Chloromonas*, *Chlainomonas*, and *Sanguina*. Although most ASVs were not assigned to genus level, most ASVs clustered near one of these genera in NMDS ordination (Figure 2). Several OTUs were dominated by ASVs that were not assigned to genus level, including three OTUs closely related to *Chloromonas*, and one Chlamydomonadaceae OTU that was only assigned to the family level.

The composition of 18S and *rbcL* libraries varied with elevation (Figure 3; Figure 4). *Sanguina* was absent below 1500 m in both 18S and *rbcL*. High elevation samples contained an unannotated OTU of *Chloromonas* that was absent from low elevation sites (Figure 3, "Chloromonas C"). Additionally, the three highest elevation sites uniquely contained ASVs with a best BLAST match to *Raphidonema longiseta* (KM462868.1)---interestingly, all of these samples were taken from snow overlying glacier. The two high alpine green snow samples were dominated by two different OTUs of *Chloromonas*, which were also both found in sites below treeline. *Chlainomonas* was prevalent at all elevations, and *Chloromonas krienitzii* was predominant in our low elevation sites.

We observed snow algae blooms dominated by *Chloromonas krienitzii* transition from green to orange from May to June (Figure 5). At two sites we observed snow melt channels or runnels containing high concentrations of snow algae. In May and early June these sites contained green snow hidden 2 to 5 cm below the white snow surface, but upon subsequent visits the same location contained orange snow at the surface, whose morphologies resembled published images of the *Chloromonas krienitzii* lineage [@matsuzaki_revisiting_2015]. We compared 18S and *rbcL* metabarcode profiles of green subsurface and orange surface snow, and found both contained high relative abundance of *Chloromonas krienitzii*. Although we did not detect *Chlainomonas* in cell counts, the surface samples contained nearly double the relative abundance of *Chlainomonas* from the subsurface samples.

<!-- include faiths diversity in supps??? thinking no  -->

@@

Here we present multiple data sets demonstrating elevational patterns in alpine snow algae blooms. *Sanguina* was dominant in red snow above treeline, while green and orange blooms of *Chloromonas krienitzii* were dominant in mid-elevation snow runnels. We found unexpected diversity within *Chloromonas* and *Chlainomonas* *rbcL* that we did not detect using 18S, including many potentially novel OTUs. In contrast to previous studies [@novis_two_2008;@prochazkova_ecophysiological_2018], our work suggests that *Chlainomonas* is widespread and abundant in alpine snow algae blooms.

Our results show bloom composition can be highly variable on the same mountain, even in nearby patches. In contrast, previous studies found snow algae assemblages were homogeneous across geographically distant sites [@lutz_biogeography_2016;@segawa_bipolar_2018]. The aforementioned studies exclusively sampled high-latitude red snow, while our study included mid-latitude red, orange, and green snow from above and below treeline. Another recent study using 18S metabarcoding found different snow algae communities in red and green snow algae blooms in Antarctica, including yellow blooms dominated by Chrysophyceae [@soto_microbial_2020]. Much of the diversity in our samples within *Chlainomonas* and *Chloromonas* was only detected with *rbcL* and not 18S, so the diversity within these genera may have been overlooked in previous metabarcoding studies.

Our *rbcL* data suggest *Chlainomonas* may be more widely distributed than previously thought. We could not differentiate between *Chlainomonas* and *Chloromonas* in our 18S data set, so this genus may have been overlooked in previous studies. *Chlainomonas* was previously thought to be restricted to waterlogged snow overlying mountain lakes [@novis_two_2008;@prochazkova_ecophysiological_2018], but we did not find any evidence that *Chlainomonas* was restricted in its habitat distribution. To the contrary, only one *Chlainomonas* dominant sample was located in waterlogged snow near the edge of a pool (sample S9); the other *Chlainomonas* dominant sites were not notably wetter than the surrounding snow, nor overlying frozen bodies of water. Interestingly, we found *Chlainomonas* was in higher relative abundance in our *rbcL* dataset than in cell counts (Figure 3)---possibly due to multiple chloroplasts per cell, which have been documented in this genus [@prochazkova_ecophysiological_2018].

Our findings highlight the remaining unexplored diversity in the snow algae microbiome. Many *rbcL* OTUs did not closely match any GenBank nucleotide archive sequences (Figure 2). "Chlamydomonadaceae E" highest BLAST percent identity was only 87%, and the top 10 BLAST matches contained 6 different algae genera---including two *Chloromonas* snow algae from Japan. This OTU did not correlate with any categories in our 18S or cell count data. One possible explanation is that it was lumped together with *Chloromonas* in 18S, and our crude quantification and categorization of cell counts did not detect this taxa. Other OTUs were also poorly matched to GenBank,  which could represent novel species of *Chloromonas* (Figure 2). The lack of clustering we observed in the remaining *Chloromonas* could be due to overlap in the levels of interspecific and intraspecific variation. *rbcL* diversity is likely higher within *Chloromonas* than in other algae genera: unlike *Sanguina* [@prochazkova_sanguina_2019], most species of *Chloromonas* lack a pyrenoid [@nozaki_differences_2002], which houses high concentrations of cross-linked Rubisco. Likely related to its lack of pyrenoid, *Chloromonas* has many non-synonymous mutations in the region of *rbcL* the codes for binding Rubisco together [@nozaki_differences_2002]. While *rbcL* is problematic in *Chloromonas* phylogeny [@nozaki_differences_2002], it nonetheless is highly differentiated and therefore an effective barcode for microalgae [@hall_assessment_2010].

<!-- *Sanguina* was found in the same samples, but relative abundance was generally higher in 18S and cell counts (Figure 3). Our *rbcL* primers were designed based on *Chloromonas* and *Chlainomonas* snow algae sequences (*Sanguina* sequence data was not available at the time), so primer bias could have resulted in over-representation of *Chloromonas* relative to *Sanguina*. -->

>>>>

---
title: "Alpine snow algae species composition along an elevational gradient"
author:
  - name: Casey B. Engstrom
    institute: SFU
  - name: Kurt M. Yakimovich
    institute: SFU
  - name: Lynne M. Quarmby
    institute: SFU
    email: quarmby@sfu.ca
institute:
  - SFU: Dept. Molecular Biology and Biochemistry, Simon Fraser University, 8888 University Drive, Burnaby, BC, V5A 1S6, Canada
output: 
  bookdown::word_document2:
    fig_caption: yes
    pandoc_args:
      - '--lua-filter=templates/scholarly-metadata.lua'
      - '--lua-filter=templates/author-info-blocks.lua'
    reference_docx: templates/word-template.docx
  # pdf_document:
  #   fig_caption: yes
  #   number_sections: yes
csl: templates/the-isme-journal.csl
bibliography: refs.bib
---

@@

The authors declare no conflicts of interest.

@@

Snow algae blooms cover vast areas of summer snowfields worldwide, and can reduce albedo and increase snow melt. Despite their global prevalence, little is known about the algae species that comprise snow algae blooms. We used 18S and *rbcL* metabarcoding and light microscopy to characterize species composition of 33 snow algae samples from alpine and subalpine habitats in the Coast Range of British Columbia, Canada. We found blooms were dominated by the snow algae genera *Sanguina*, *Chloromonas*, and *Chlainomonas*. This study is the first to thoroughly document regional variation between blooms, and we found that blooms were highly variable in species composition, with high elevation sites containing distinct species assemblages from forested sites. In contrast to previous work, the snow algae genus *Chlainomonas* was an abundant and widespread in snow algae blooms. Although we found few taxa using traditional 18S metabarcoding, *rbcL* revealed diversity with high taxonomic resolution, including OTUs that could represent novel species of snow algae. These three cross-referenced datasets show alpine snow algae blooms are more diverse than previously thought, and are highly variable from site to site along an elevational gradient.

@@

# Introduction {-}
<!-- global, vast blooms. albedo->snowmelt. what species? -->
Each summer, vast areas of snow surface are coloured red by snow algae blooms in polar and alpine snowfields worldwide. Red snowfields have been found on every continent [@segawa_bipolar_2018;@duval_snow_1999;@marchant_snow_1982;@yoshimura_community_1997;@vimercati_nieves_2019] and overlying Arctic sea ice [@gradinger_snow_1996]. The distribution and extent of snow algae blooms are poorly understood, but they can be quite extensive: in Alaska, remote sensing detected snow algae in one third of a 1,900 km^2^ icefield [@ganey_role_2017]. In recent years snow algae have received attention for their role in reducing snow surface albedo, which could substantially impact the rate of snow melt [@lutz_biogeography_2016;@ganey_role_2017]. Potentially, snow algae could substantially impact the timing of spring melt, which in turn could deplete summer water supplies in mountain snowpack and reduce glacier mass balance. Snow algae blooms have been recorded throughout history since the time of the ancient Romans [@mayhoff_naturalis_1906;@darwin_voyage_1959], yet despite their global prevalence we are only beginning to identify the species that cause snow algae blooms.

Microscopy reveals a diversity of cell morphologies in snow algae blooms, but different species can look nearly identical [@matsuzaki_taxonomic_2019], and the same species can look completely different under different environmental conditions. The snow algae *Chloromonas krienitzii* are small green biflagellates in culture, but in field samples cells of this species are nearly twice the diameter, orange-pigmented, with thick cell walls and short spines [@matsuzaki_revisiting_2015]. The environmental cues that trigger this transformation are not well understood, but in the freshwater green algae *Haematococcus pluvialis* red secondary pigments are produced in response to high light [@shah_astaxanthin-producing_2016]. Green snow algae blooms are less frequently reported, and microscopy-based studies have suggested that green snow develops into red snow [@muller_persistent_2001]. However, metabarcoding studies to date have found that green snow and red snow contain different species [@terashima_microbial_2017;@lutz_integrated_2015].

@@

<!-- metabarcoding, red snow looks the same -->
Although there are many species of snow algae, only a few OTUs seem to dominate red snow algae blooms worldwide. 33 Arctic red snow samples were all dominated by "uncultured Chlamydomonadaceae 1 and 2" OTUs with low relative abundance of *Chloromonas polyptera* and *Raphidonema nivale* [@lutz_biogeography_2016]. Similarly, another study using metabarcoding found 24 red snow sites across the Arctic and Antarctic contained similar assemblages, dominated by "uncultured Chlamydomonadaceae A and B" with secondary abundance of *Raphidonema* and *Chloromonadinia* [@segawa_bipolar_2018]. Other 18S metabarcoding studies of snow algae were limited to genus or family level taxonomy [@hamilton_primary_2017]---because 18S is highly conserved, this marker cannot distinguish between closely related species or genera.

<!-- our goal, what we did, what we found -->
Virtually nothing is known about the regional variation in species composition of snow algae blooms, particularly in alpine regions. We set out to identify the algae in blooms on different mountains and elevations in the Coast Range of British Columbia, Canada. We assessed species composition using light microscopy and 18S and *rbcL* metabarcoding. *rbcL* codes for the large subunit of Rubisco, thus targeting only the photosynthetic component of the microbiome. Using these three cross-referenced metrics we describe the algal species assemblage in our study area with high accuracy. *rbcL* revealed much greater diversity than would have been found using 18S alone. We did not find any differences in bloom composition between mountains, but there were distinct algae species assemblages at different elevations.

@@

@@

We collected 309 snow algae samples over the summer of 2018 from alpine and subalpine sites in the Coast Range near Vancouver, British Columbia, Canada (Supplementary Figure S1). We collected red, orange, and green snow samples from 13 different mountains from sites ranging from 880 to 2150 m above sea level (Supplementary Table S1). Over the summer we collected samples from progressively higher elevations as snow melted at lower elevations, and we did not detect snow algae at higher elevations until later in the summer. We scooped samples from visibly coloured snow into sterile 50 mL centrifuge tubes. We kept samples cold during transport back to the lab by packing tubes in snow. We melted each sample at room temperature on the bench and removed a 1 mL aliquot for light microscopy. Samples were then stored at -20 &deg;C for up to eight months until DNA extraction.

We used light microscopy to characterize cell morphology within 24 hours after collection. We fixed samples in 2% gluteraldehyde, and under 400 x magnification observed the 100 cells closest to the center of the slide. We classified cell morphology based on similarity to published photographs of *Sanguina nivaloides* [@prochazkova_sanguina_2019], *Chloromonas cf. nivalis* [@prochazkova_chloromonas_2018], *Chloromonas cf. brevispina* [@matsuzaki_revisiting_2015], *Chloromonas krienitzii* [@matsuzaki_revisiting_2015], *Chlainomonas rubra* [@novis_two_2008], and for cells that did not fall into one of there categories as either "green cell" or "other".

<!-- collected a ton of samples, characterized morphology -->

@@

<!-- arising implicit hypothesis 1) different morph species, different genetic species 2) different elevation, different species (our methods not ideal for this, should have been randomized sample if we wanted to test for elevation) -->

We Illumina sequenced *rbcL* and 18S amplicons of 33 samples. We selected this subset for sequencing to compare the species composition of blooms containing distinct morphologies, and to compare the genetic composition of morphologically similar blooms from different mountains, elevations, and dates. We freeze-dried these samples for up to 48 hours until until samples appeared completely desiccated. We mini-pestled between 5 to 20 mg of sample at room temperature to physically rupture cell walls before incubation in CTAB lysis buffer [@noauthor_ctab_2009]. We extracted DNA in batches of 5-6 samples using chloroform:isoamyl alcohol [@cubero_dna_1999], and purified DNA using ethanol and spin columns (Qiagen, Hilden) (Supplementary Protocol S1). As a negative control against cross-contamination we processed a tube of sterile distilled water alongside each batch, and tested this for DNA with Qubit fluorometric quantification (Thermo Fisher, Waltham, MA).

In addition to 18S, we designed custom snow algae *rbcL* primers. We chose this marker because *rbcL* is more differentiated between snow algae than 18S (Supplementary Figure S2), and there is snow algae *rbcL* reference data on GenBank. Our primers were designed to target a 400 bp section of snow algae *rbcL*, based off the consensus of 20 GenBank snow algae sequences (Supplementary Figure S2). *Sanguina* sequences were not included because they were not available at the time. We designed our primers with the Eurofins primer design tool (https://www.eurofinsgenomics.eu/en/ecom/tools/pcr-primer-design/). To target 18S, we used the universal primers Euk1181 and Euk1624 targeting the V7-V8 hypervariable regions [@wang_optimal_2014]. Primer sequences are available in Supplementary Table S2.

We constructed our 18S and *rbcL* amplicon libraries using a standard two-step PCR approach [@meyer_illumina_2010] (Supplementary Protocol S2). After each PCR, we purified using Agencourt AMPure XP kit (Beckman Coulter, Brea, CA). We standardized sample DNA concentration with Qubit, pooled our library, and sequenced this on a MiSeq using the V3 kit following manufacturer instructions (Illumina, San Diego, CA).

@@

We demultiplexed reads with CUTADAPT [@martin_cutadapt_2011], and filtered, trimmed, denoised, dereplicated, merged paired-end reads, and removed chimeras following the default pipeline of DADA2 [@callahan_dada2_2016]. We assigned taxonomy to amplicon sequence variants (ASVs) using IDTaxa [@murali_idtaxa_2018] using a high confidence threshold parameter (t=50). Because snow algae are not well represented on curated databases such as SILVA, we made custom databases using GenBank snow algae sequences (scripts and databases available on github). To ensure that we were not overlooking non-snow algae taxa we also ran our 18S library against SILVA [@quast_silva_2013].

@@

<!-- ??? make sure this is on GH: -->

@@

To assign OTUs, we clustered ASVs using t-SNE [@maaten_visualizing_2008] and DBSCAN [@hahsler_dbscan_2019]. To ensure that our t-SNE was not an artifact of parameter selection, we ran t-SNE with different values of "perplexity", and found the same clusters were generated from 10 to 50. To cross-validate this unconventional OTU clustering method, we overlaid these OTU clusters on phylogenetic trees using IQTree [@nguyen_iq-tree:_2015] (Supplementary Figure S3).

<!-- mention swarm? would need to show this dint work on tree as supp, no time -->

# Results {-}

<!-- add read depth, etc? -->
<!-- i think the cell counts needs to go into the main text -->

We observed morphologically distinct snow algae blooms at different elevations (Figure 1; Supplementary Figure S4). Red snow was prevalent in areas of high solar exposure above treeline, predominantly containing cells similar in appearance to *Sanguina nivaloides* [@prochazkova_sanguina_2019]. We found cells resembling *Chlainomonas rubra* [@novis_two_2008] across all elevations. Green snow patches were infrequent in alpine meadows receiving full sun, and more common in shady forested sites. Forested sites often contained green or orange patches of snow, many of these sites contained green or orange ovate cells with short spines resembling *Chloromonas cf. brevispina* [@matsuzaki_revisiting_2015].

Both 18S and *rbcL* amplicon libraries were dominated reads assigned to Chlorophyta (Figure 2). Our 18S library assigned with SILVA contained 50 Chlorophyceae, 11 Trebouxiophyceae, and 7 Chrysophyceae ASVs, while our *rbcL* library contained 603 Chlorophyta and 41 Trebouxiophyceae ASVs. Both ASV and OTU diversity was higher in *rbcL* than 18S: nine *rbcL* OTUs, compared to just three in 18S (Supplementary Figure S5). The most frequent *rbcL* genera were *Chloromonas*, *Chlainomonas*, and *Sanguina*. Although the majority of ASVs were not assigned to genus level, most ASVs were genetically similar to one of these three genera, with one exception. Sequences in the OTU E were not assigned to genus level; the ten best BLAST matches included six different genera within Chlamydomonadaceae (86-87% sequence match), two of which were *Chloromonas* snow algae (LC012735). Many ASVs were similar to *Chloromonas*, including two distinct OTU clusters D and F which did not match any known species within this genus.

18S and *rbcL* taxonomic composition varied with elevation. Low elevation samples clustered together in *rbcL* UniFrac [@lozupone_unifrac_2005], forming a cluster that was distinct from most high elevation samples (Figure 3). We found the highest diversity in samples that were collected latest in the season, although we did not find a statistically significant trend between Shannon diversity and date there was a weak correlation between Faith's PD and date (Pearson's *r*=0.36, p=0.04) (Supplementary Figure S6). In both 18S and *rbcL*, *Sanguina* predominated many sites above 1500 m, but was absent below this elevation (Figure 4). Additionally, high elevation samples contained one OTU of *Chloromonas* that was absent from low elevation sites (Figure 3, OTU F). The three highest elevation sites uniquely contained ASVs with a best BLAST match to *Raphidonema longiseta* (KM462868.1)---interestingly, all of these samples were taken from snow overlying glacier. The two high alpine green snow samples were dominated by two different OTUs of *Chloromonas*, which were also both found in sites below treeline. *Chlainomonas* was prevalent at all elevations, and *Chloromonas krienitzii* was predominant in our low elevation sites.

A green to orange transition occurred from May to June at two sites dominated by *Chloromonas krienitzii*. On initial visits to a site in May we observed green snow hidden 2 to 5 cm below the snow surface (sample S2.2), but on subsequent visits in June (samples S4.6, S6.1, S9.2, S11.2) the same runnel was orange on the surface overlying the green snow. Microscopy of orange snow revealed orange spherical cells with short spines resembling *Chloromonas krienitzii* [@matsuzaki_revisiting_2015], while green snow contained motile green flagellates. *rbcL* comparison of green and orange snow showed both were dominated by reads assigned to *Chloromonas krienitzii*, with the orange snow containing higher abundance of *Chlainomonas* (Supplementary Figure S7).

@@

Here we present multiple data sets demonstrating elevational patterns in alpine snow algae blooms. *Sanguina* was dominant in red snow above treeline, while green and orange blooms of *Chloromonas krienitzii* were dominant in runnels at lower elevations. We found unexpected diversity within *Chloromonas* and *Chlainomonas* *rbcL* that we did not detect using 18S, including many potentially novel OTUs. Previously only reported from red snow overlying lakes [@prochazkova_ecophysiological_2018], *Chlainomonas* was widespread and abundant in alpine snow algae blooms.

Our results show bloom composition can be highly variable on the same mountain, even in nearby patches. In contrast, previous studies found red snow algae assemblages were homogeneous across geographically distant polar sites [@lutz_biogeography_2016;@segawa_bipolar_2018]. This discrepancy could be because our study included green and orange snow samples, or it could be that mid-latitude mountains have higher snow algae diversity than polar sites. We noted most snow algae blooms above treeline were dominated by *Sanguina*; by only collecting a single sample as representative of a region could have meant that the previous studies could have overlooked less common snow algae communities. We detected more OTUs with *rbcL* than 18S, so previous 18S metabarcoding could have overlooked snow algae diversity at lower taxonomic levels.

Our *rbcL* data suggest *Chlainomonas* may be more widely distributed than previously thought. *Chlainomonas* was previously thought to be restricted to waterlogged snow overlying mountain lakes [@novis_two_2008;@prochazkova_ecophysiological_2018], but we did not find any evidence that *Chlainomonas* was restricted in its habitat distribution. To the contrary, only one *Chlainomonas* dominant sample was located in waterlogged snow near the edge of a pool (sample S9), while the other *Chlainomonas* dominant sites were not notably wetter than the surrounding snow, nor were they located over frozen lakes. We could not differentiate between *Chlainomonas* and *Chloromonas* in our 18S data set, so this genus may have remained undetected in previous 18S metabarcoding. Interestingly, *Chlainomonas* was in higher relative abundance in our *rbcL* dataset than in cell counts (Figure 3). One possible explanation is high *rbcL* copy number: the large cells of *Chlainomonas rubra* have multiple parietal chloroplasts per cell [@prochazkova_ecophysiological_2018], and *rbcL* is located in the plastid genome.

Our findings highlight the remaining unexplored diversity in the snow algae microbiome. Several *rbcL* OTUs did not closely match any GenBank nucleotide archive sequences, such as the OTU "Chlamydomonadaceae E" (Figure 2). This OTU was distinct in our *rbcL* data, but it did not correlate with any categories in 18S or cell count data. One possible explanation is that it was lumped together with *Chloromonas* in 18S, and our crude quantification and categorization of cell counts did not detect this taxa. OTUs D and F were related to *Chloromonas*, but contained few ASVs that were assigned to genus (Figure 2). Possibly these OTUs represent novel species of *Chloromonas*. However, there were no discernible clusters within the OTU "Other *Chloromonas*", perhaps due to overlap in the levels of interspecific and intraspecific variation. *rbcL* diversity is likely higher within *Chloromonas* than in other algae genera: unlike *Sanguina* [@prochazkova_sanguina_2019], most species of *Chloromonas* lack a pyrenoid [@nozaki_differences_2002], which houses high concentrations of cross-linked Rubisco. Lacking pyrenoid, *Chloromonas* have many non-synonymous mutations in the region of *rbcL* the codes for binding Rubisco together [@nozaki_differences_2002]. While *rbcL* is a poor indicator of *Chloromonas* phylogeny [@nozaki_differences_2002], it nonetheless is highly differentiated and therefore an effective barcode for microalgae [@hall_assessment_2010].

Variation in bloom composition could be due to a wide range of habitat features. *Sanguina* and *Chloromonas* C were limited to sites above 1500 m in full sunlight, but many low elevation sites also received full sunlight and did not contain these OTUs (Figure 3). Light intensity, snow moisture, or snow chemistry could all plausibly influence community composition. Intriguingly, we only observed *Raphidonema* at high-elevation glacier sites. In Svalbard, *Raphidonema nivale* abundance increased on glacier surface snow following wind storms, and the authors concluded that *Raphidonema* was a soil algae that opportunistically colonizes snow following wind deposition, but being sub-optimally adapted for snow such blooms are short lived [@stibal_growth_2005]. Given the aerial dispersal capabilities of microalgae [@tesson_airborne_2016] and genetic overlap between distant snow algae populations [@segawa_bipolar_2018] it seems unlikely that geographic distance is a barrier to snow algae distribution. Underlying topography could potentially influence species distribution: two sites dominated by *Chloromonas krienitzii* were in runnels overlying ephemeral streams, which could influence nutrient availability where surface debris is deposited in the runnel.

<!-- As i presented it above, looks like Chlainomonas is causing the bloom. Need to explain the nuance -->
Previous work has shown that *Chloromonas krienitzii* undergoes distinct green and orange morphologies [@matsuzaki_revisiting_2015], but this paper is the first to document this transition in the field. The transition over weeks suggests that this process is mediated by seasonal changes. Secondary pigments likely protect snow algae from the damaging effects of intense solar irradiation at the snow surface [@bidigare_evidence_1993], which could be why green cells were most concentrated a few centimeters below the snow surface. *Chlainomonas* was more abundant in surface samples as measured by *rbcL*, but we microscopy suggested that *Chlainomonas* was only present in low relative abundance.
